Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2023-10-21 go3.tex
sent for comment
submitted version

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Beyond compositionally in high throughput sequencing; estimating the importance of scale in data analysis with ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Beyond compositionally in high throughput sequencing; estimating
the importance of scale in data analysis with ALDEx2}
\author{Gregory B. Gloor$^1$ \and Michelle Pistner Nixon$^2$ \and Justin Silverman$^{2,3}$}
\date{\today}

\begin{document}
\maketitle

$^1$ Department of Biochemistry, University of Western Ontario
$^2$ College of Information Sciences and Technology, Pennsylvania  State University
$^3$ Department of Medicine, Pennsylvania  State University

\begin{abstract}
Despite scale being an important confounding variable, high throughput
sequencing generates data that are difficult to reconcile with the scale
of the underlying environment, but it has been shown that the scale of
the environment is related to the geometric mean of the parts of the
sample. The ALDEx2 R package builds a Bayesian model of the data and
here we report an update to the model that includes both random sampling
and scale estimates. We show that incorporating scale allows for robust
inference in a number of different experimental designs and we provide
guidance on how to incorporate scale inference into differential
abundance analysis. While all data normalizations in widespread use make
implicit assumptions about the scale of the system we suggest that
explicit rather than implicit scale models should be used.
\end{abstract}

@@

As shown using the effect (Gloor, Macklaim, and Fernandes 2016) and
volcano plots (Cui and Churchill 2003) in Figure 2 A,B,D,E the root
cause of the many statistically significant positive transcripts is the
very large number of transcripts with negligible variance with both
DESeq2 and ALDEx2. We can see that almost all the transcripts that are
differentially abundant with an FDR \textless{} 0.01 (orange and red
points) have extremely low dispersion and a very low difference between
groups. In the most extreme cases transcripts with near 0 difference
have a low FDR. This issue lead to the common practice of choosing
transcripts with a low FDR and a fold-change threshold (Schurch et al.
2016) (commonly set at \(\pm 2^{1.4}\)), and these limits are shown by
the dashed grey lines. A similar sitation arises when using the ALDEx2
package, and indeed the two methods identify substantially similar
transcripts. This results begs the question, ``why bother with the
signficance test?''.

The very low dispersion estimate for most of the features arises because
scale variation in the underlying dataset has been removed through
sequencing and normalization. While the actual scale of the data is
inaccessible post-sequencing we can estimate the \emph{effect} of scale
on the output. To do this, we add uncertainty to the denominator that is
used to calculate the log-ratio of the samples (Nixon et al. 2023), and
combine this with the posterior probabilities that ALDEx2 calculates on
a per-feature basis (Fernandes et al. 2013). Scale uncertainty is
incorporated using the \texttt{gamma} parameter that controls the amount
uncertainty being included when we call either \texttt{aldex()}, or
\texttt{aldex.clr()}. The \texttt{ALDEx2} package further contains a
sensitivity analysis function, \texttt{aldex.senAnalysis()}, that can be
used to explore the effect of different amounts of scale uncertainty. In
practice we suggest that a \texttt{gamma} parameter between 0.5 and 1 is
realistic for most experimental designs.

Applying \texttt{gamma=1} as a parameter we can see that the large
number of transcripts with near 0 dispersion have had their dispersion
increased (Figure 2C), and this results in many fewer transcripts (100)
being called significantly different as shown in the volcano plot in
Figure 1F (red points). Furthermore, overplotting the significant
transcripts identified after adding scale uncertainty on the un-scaled
analysis shows that adding scale uncertainty removes the need for the
dual cutoff. Indeed, adding scale uncertainty reduces the significant
transcripts to approximately the number observed with the somewhat
arbitrary difference cutoff. Thus, incorporating scale uncertainty
through the default scale model allows us to determine which variables
are likely to be significant due to sequencing and normalization, and
which are significantly different even with scale uncertainty included.

The second example dataset is a vaginal metatranscriptome dataset used
in (Macklaim and Gloor 2018; Wu et al. 2021), where we are comparing
gene expression in bacteria collected from healthy (H) and bacterial
vaginosis (BV) affected women. In this environment, both the relative
abundance of species between groups is different as is the gene
expression levels within a species (Macklaim et al. 2013). We further
know that the total number of bacteria is about 10X more in BV than in H
(Zozaya-Hinchliffe et al. 2010). Thus, this is an extremely challenging
environment in which to determine differential abundance. Indeed, the
accepted method to not analyze the vaginal metatranscriptome as a whole
system but instead to do the simpler taxon-by-taxon (Macklaim et al.
2013; Deng et al. 2018; Fettweis et al. 2019). The problem with the
systemic analysis is that the scale confounding results in many
housekeeping genes as being differentially abundant between groups.

In this example we show how to specify the scale model explicitly and
demonstrate that applying different scale models to each group
independently can control for the very large difference in scale in the
underlying data. When specifying the whole scale model we can pass a
matrix of scale values instead of a single estimate of gamma to
\texttt{aldex.clr()}. This matrix should have the same number of rows as
the of Monte-Carlo Dirichlet samples, and the same number of columns as
the number of samples. There is an \texttt{aldex.makeScaleModel()}
function that can be used to simplify this step.

This scale model matrix encapsulates two aspects of scale uncertainty
and explicitly illustrates the concept of scale. First, as in the first
example, the gamma paramter indicates the scale uncertainty. Second, we
need to explicitly indicate the difference in scale between the groups
in the dataset.

Figure 3A shows an effect plot (Gloor, Macklaim, and Fernandes 2016) of
the data where reads are grouped by function, corresponding
approximately to grouping orthologous sequences regardless of the
organism of origin. Each point represents one of the 3728 functions, and
we can see that there are many more functions represented in the BV
group (bottom) than in the healthy group (top). This is because the
\textit{Lactobacilli} that dominate a healthy vaginal microbiome have
reduced genome content relative to the anaerobic organisms that dominate
in BV, because there is a greater diversity of organisms in BV than in H
samples and because the BV condition has at least an order of magnitude
more bacteria than does the H condition.

We can see that there are a large number of functions that are shared
between the two groups (boxed area in Figure 3A), and inspection shows
that these largely correspond to core metabolic functions that would not
be expected to contribute to differences in ecosystem behaviour. As a
proxy for housekeeping functions the core ribosome functions (blue)
shows that their mean location is not centred on 0. The major group of
these housekeeping functions is located off the line of no difference
(being approximately located at +1.5) and not surprisingly have among
the lowest dispersion in the dataset. Nevertheless, they are identified
as differentially abundant (red) along with many others. While changes
in the abundance of housekeeping functions is a useful proxy for
relative abundance of species in the environment, they tell us nothing
about the functional capacity of the two groups because these are
functions in common to every organism. Of more interest is determining
the functions that are different between groups because these are unique
or over-expressed in one group relative to the other.

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/meta-1.pdf}
\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
Orthology (KO) functional level. Panel A shows an effect plot for the
default analysis where the functions that are elevated in the healthy
individuals have positive values and functions that are elevated in BV
have negative values. Highlighed in the box are highly abundant KOs that
are almost exlusively housekeeping functions, with ribosomal KOs
highlighted in blue, statistically significant (FDR \textless{} 0.01)
functions in red, and non-significant functions in black or orange.
These housekeeping functions should be located on the midline of no
difference. Panel B shows the same data scaled with
\texttt{gamma\ =\ 1}, which increase the minimum dispersion approximatly
by one unit. Here the housekeeping functions from Panel A are colored
cyan or blue for reference. Panel C shows the same data scaled with
\texttt{gamma\ =\ 0.75} and a 0.15 fold difference in dispersion applied
to the BV samples relative to the H samples. The orange functions are
now statistically significant. Note that this shifts the midpoint of the
housekeeping functions towards the midline.}
\end{figure}

Applying the default scale model of \texttt{gamma=1} as before increases
the dispersion as expected, but does little to move the large number of
housekeeping functions toward the midline of no difference.
Nevertheless, about 50\% of the housekeeping functions are no longer
statistically significantly different.

Up to this point, scale uncertainty has been applied uniformly to both
conditions, but a user-defined scale adjustment can be applied to each
condition, or even each sample independently through a custom scale
matrix. For this we need to revisit the concept that all normalizations
in widespread use are actually ratios with the denominator chosen by one
of several methods. It is important to recognize that in some datasets
the denominator itself is not crucial as all approaches give
substantially similar results; the transcriptome example used previously
is a good example of such an idealized dataset. This begs the question,
``if it is not the denominator, what then is the important
determinant?'' As shown by Nixon et al. (2023) it is usually the
\emph{relationship} between the group scale values (denominators) that
is important, not their actual value. This can be illustrated quite
simply by starting with the mean ratio in \(G\) between groups for the
yeast transcriptome dataset which are \(6.05\times^{-5}\) for snf-w and
\(5.08\times^{-5}\); their ratio being 1.17, or a 0.17-fold difference.
Using this information, we can recapitulate the differential abundance
analysis in Figure 2B and 2C exactly by using setting the mean
denominator of group 1 to 1, and group 2 to 1.17 with a gamma of 1 as
shown in Supplemental Figure 2. Altering this ratio moves the location
of the data.

With this background, we can understand how to apply a full scale to the
metatranscriptome dataset with the \texttt{aldex.makeScaleMatrix()}
function. This function uses a logNormal distribution when building the
model so that ratios that are chosen are symmetrical. The full scale
model option can be used to make a matrix of relative scales that are
distinct for each group (or even each sample) along with their
uncertainties. Applying a per-group relative differential scale of 0.15
with a base scale of 1 moves the housekeeping functions to the midline
of no difference (Figure 3C). Note that now a significant number of
functions are differentially up in BV that were formerly classed as not
different without scale, or when only a uniform scale was applied. These
former false negatives are noted in orange in each panel. Inspection of
the functions shows that these are largely missing from the
Lactobacillus species and so should actually be captured as
differentially abundant. Thus, applying differential scale allows us to
distinguish between both false positives (housekeeping functions in
cyan) and false negatives (orange functions) even in a very difficult to
analyze dataset. We suggest that the default scale model alone is useful
and should be normally applied when the data are well centred. However,
when datasets are not well centred or when the investigator has prior
information about the underlying biology, we advocate developing and
using a full differential scale model.

@@

Biological count data derived from HTS can be decomposed into two parts
the relative (compositional) and the absolute (scale), and the product
of these generates a fully scaled biological system (Nixon et al. 2023).
Biological systems are both predictably variable and stochastic and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation. In the absence of information external
to the sequencing run itself, no normalisation method can recapture any
of the scale information (Lovén et al. 2012).

Since the underlying scale of the system cannot be measured easily --
methods such as flow cytometry (Vandeputte et al. 2017), spike-in probes
and fluorescent in-situ hybridization (Lovén et al. 2012; Marguerat et
al. 2012) have been used -- the effect on analysis can be modelled by
including scale uncertainty in the analysis (Nixon et al. 2023). Nixon
et al, showed that this can be done by including uncertainty in the
denominator used for the normalization. The ALDEx2 R package is ideally
suited for this since this tool builds a Bayesian posterior of the
compositional component of the dataset at the outset and then conducts
the analysis on that posterior. Adding scale uncertainty can be done at
the same time thus producing a posterior model that incorporates both
compositional and scale uncertainty, and this can be done without
additional computational complexity. For this, the compositional
uncertainty is sampled from a Dirichlet distribution, and the scale
uncertainty can be sampled from any distribution depending on prior
knowledge or preference, but ALDEx2 samples from a logNormal
distribution by default.

All normalizations attempt to make the samples in a dataset commensurate
but do not explicitly address the scale of the underlying system. The
two supplementary tables show that these normalizations can be
interpreted as scale assumptions. However, the general lack of scale
information has important consequences for the analysis of HTS datasets.
One issue is that analysis tools seem over-powered with even moderate
sample sizes and different tools have different intrinsic power, Type 1
and Type 2 error rates (Schurch et al. 2016). Using small sample sizes
in analysis leads to less reliability and reproducibility in analyses
since surprisingly large sample sizes are needed to determine
reproducible p-values (eg. (Halsey et al. 2015)). Thus, recommendations
to use small sample sizes in multivariate datasets such as RNA-seq
datasets are not supported by simple modelling in the univariate case.
Another issue is that datasets are difficult to analyze when there they
contain systematic asymmetry, with different tools exhibiting differing
pathologies with these datasets (Robinson and Oshlack 2010; Wu et al.
2021).

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as transcriptomes
(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
Quinn, Crowley, and Richardson 2018) and microbiomes(Thorsen et al.
2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
observed that the actual agreement between methods can be modest, and
there is usually a recommendation that some methods are better suited
for individual datasets or for specific types of analysis;
e.g.~`Normalization and microbial differential abundance strategies
depend upon data characteristics' (Weiss et al. 2017). However, another
interpretation of these studies is that the specific assumptions of the
tools and the way they are used by different groups can lead to results
that include both false negative and false positive results, skewed
towards the latter. There are now published examples of multiple large
replication efforts in psychology (Open Science Collaboration 2015),
cancer biology (Rodgers and Collings 2021), ecology and evolution (Gould
and et.al, n.d.). In a nutshell, experiences from these studies indicate
that published data typically have higher effect sizes and appear more
significant than replication efforts with the same data, or with newly
generated replication data (Rodgers and Collings 2021). Thus, most
published data should be viewed with suspicion (Ioannidis 2005) unless
efforts have been made to ensure that the results are robust

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Normalizations are not what you think; Explicit Scale Simulation in ALDEx2},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Normalizations are not what you think; Explicit Scale Simulation
in ALDEx2}
\author{true \and true \and true}
\date{}

\begin{document}
\maketitle
\begin{abstract}
In many high-throughput sequencing (HTS) studies, sample-to-sample
variation in sequencing depth is not driven by variation in the scale
(e.g., total size, microbial load, or total gene expression) of the
underlying biological systems being measured but rather is driven by
technical factors. Typically the technical variation is addressed using
some form of statistical normalization. Normalizations are data or
parameter transformations that remove unwanted technical variation in
the hopes of facilitating analyses sensitive to scale; e.g.,
differential abundance and differential expression analyses. Recently we
showed that any normalization makes implicit assumptions about the
unmeasured system scale and that errors in these assumptions can lead to
dramatic increases in false positive and false negative rates. Here we
describe updates to the ALDEx2 R package that mitigate these problems by
directly modeling uncertainty in the unmeasured system scale through the
use of a \textit{scale model}. Scale models generalize the idea of
normalizations and can be thought of as explicitly modeling potential
error in the chosen normalization. Beyond enhancing the robustness of
HTS analyses, the use of scale models within ALDEx2 enhances the
transparency and reproducibility of analyses by making implicit
normalizing assumptions an explicit part of the model building process.
\end{abstract}

@@

As shown using effect (Gloor, Macklaim, and Fernandes 2016) and volcano
plots (Cui and Churchill 2003) in Figure 2 A,B,D,E, the root cause of
the many statistically significant positive transcripts is the very
large number of transcripts with negligible variance. This phenomenon is
shared between methods: almost all the transcripts that are
differentially abundant with an FDR \textless 0.01 (orange and red
points) have extremely low dispersion and a very low difference between
groups. In the most extreme cases, some transcripts have a near zero
difference between groups. This issue is not unique to this data set and
it is common practice to use a dual-cutoff by choosing transcripts based
on a thresholds for both corrected p-values and fold-changes (Schurch et
al. 2016) (commonly set at \(\pm 2^{1.4}\) for the latter). These limits
are shown by the dashed grey lines. In this case, applying a dual-cutoff
using a heuristic of at least a \(2^{1.4}\) fold change reduces the
number of significant outputs to 193 for DESeq2 and to 186 for ALDEx2.
This result begs the question: why bother with significance tests at
all?

The very low dispersion estimate for most features is a by-product of
sequencing and normalization. While the actual scale of the data is
inaccessible post-sequencing, we can study the sensitivity of the
results to the choice of normalization and thus scale. To do this we
incorporate scale models (Nixon et al. 2023) in lieu of normalizations
through the default scale model that is now built into ALDEx2. Scale
uncertainty is incorporated using the \texttt{gamma} parameter that
controls the amount uncertainty added to the CLR mean assumption when we
call either \texttt{aldex()}, or \texttt{aldex.clr()}. The
\texttt{ALDEx2} package further contains a sensitivity analysis
function, \texttt{aldex.senAnalysis()}, that can be used to explore the
effect of different amounts of scale uncertainty. In practice, we
suggest that a \texttt{gamma} parameter between 0.5 and 1 is realistic
for most experimental designs.

Applying the default scale model with \texttt{gamma=1}, a large number
of transcripts with near 0 dispersion have had their dispersion
increased (Figure 2C), and this results in many fewer transcripts 100
being called significantly different as shown in the volcano plot in
Figure 1F (red points). Furthermore, overplotting the significant
transcripts identified after adding scale uncertainty on the un-scaled
analysis shows that adding scale uncertainty removes the need for the
dual cutoff. Indeed, adding scale uncertainty reduces the significant
transcripts to a subset of those identified with the dual cutoff that
have the largest effect size. Thus, incorporating scale uncertainty
through the default scale model allows us to determine which variables
are likely to be significant due to sequencing and normalization, and
which are significantly different even with scale uncertainty included.

\hypertarget{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}{%
\subsection{Housekeeping genes can be used to guide scale model
choices.}\label{housekeeping-genes-can-be-used-to-guide-scale-model-choices.}}

Macklaim and Gloor (2018) and Wu et al. (2021) used a vaginal
metatranscriptome dataset used to compare the gene expression in
bacteria collected from healthy (H) and bacterial vaginosis (BV)
affected women. In this environment, both the relative abundance of
species between groups and the gene expression level within a species is
different (Macklaim et al. 2013). Additionally, prior research suggests
that the total number of bacteria is about 10 times more in the BV than
in the H condition (Zozaya-Hinchliffe et al. 2010). Thus, this is an
extremely challenging environment in which to determine differential
abundance as there are both compositional and scale changes between
conditions. The accepted method to analyze vaginal metratranscriptome
data is on a taxon-by-taxon basis (Macklaim et al. 2013; Deng et al.
2018; Fettweis et al. 2019) because the scale confounding can be
ignored. The problem with a systemic analysis is that many housekeeping
genes are returned as differentially abundant between groups, a result
likely due to a disconnect between the scale assumptions of the
normalization used (Wu et al. 2021) and the actual change in scale
between conditions.

In this example, we show how to specify the scale model explicitly and
demonstrate that applying a user-defined scale model can account for
some of these modeling difficulties. A user-defined scale model can
control for both the mean difference of scale between groups (e.g.,
directly incorporate information on the differences in total number of
bacteria between the BV and H conditions) as well as the uncertainty
assumed in that difference. To specify a user-defined scale model, we
can pass a matrix of scale values instead of a single estimate of gamma
to \texttt{aldex.clr()}. This matrix should have the same number of rows
as the of Monte-Carlo Dirichlet samples, and the same number of columns
as the number of samples. While this matrix can be computed from scratch
by the analyst, there is an \texttt{aldex.makeScaleModel()} function
that can be used to simplify this step in most cases.

Figure 3A shows an effect plot (Gloor, Macklaim, and Fernandes 2016) of
the data where reads are grouped by function, corresponding
approximately to grouping orthologous sequences regardless of the
organism of origin. Each point represents one of the 3728 functions.
There are many more functions represented in the BV group (bottom) than
in the healthy group (top). This is because the \textit{Lactobacilli}
that dominate a healthy vaginal microbiome have reduced genome content
relative to the anaerobic organisms that dominate in BV, because there
is a greater diversity of organisms in BV than in H samples and because
the BV condition has at least an order of magnitude more bacteria than
does the H condition.

We can see that there are a large number of core metabolic and
information processing functions that are shared between the two groups
(boxed area in Figure 3A), and inspection shows that these largely
correspond to core metabolic functions that would not be expected to
contribute to differences in ecosystem behaviour and so should not be
differentially abundant. As a proxy for housekeeping functions the core
ribosome functions (blue) shows that their mean location is not centred
on 0. The major group of these housekeeping functions is located off the
line of no difference (being approximately located at +1.5). Since they
have very low dispersions, these functions are identified as
differentially abundant (red) along with many others. While changes in
the abundance of housekeeping functions is a useful proxy for relative
abundance of species in the environment, they tell us nothing about the
functional capacity of the two groups because these are functions in
common to every organism. Of more interest is determining the functions
that are different between groups because these are unique or
over-expressed in one group relative to the other.

\begin{figure}
\centering
\includegraphics{go3_files/figure-latex/meta-1.pdf}
\caption{Analysis of vaginal transcriptome data aggregated at the Kegg
Orthology (KO) functional level. Panel A shows an effect plot for the
default analysis where the functions that are elevated in the healthy
individuals have positive values and functions that are elevated in BV
have negative values. Highlighed in the box are highly abundant KOs that
are almost exlusively housekeeping functions, with ribosomal KOs
highlighted in blue, statistically significant (FDR \textless{} 0.01)
functions in red, and non-significant functions in black or orange.
These housekeeping functions should be located on the midline of no
difference. Panel B shows the same data scaled with
\texttt{gamma\ =\ 0.5}, which increase the minimum dispersion
approximatly by one unit. Here the housekeeping functions from Panel A
are colored cyan or blue for reference. Panel C shows the same data
scaled with \texttt{gamma\ =\ 0.5} and a 0.15 fold difference in
dispersion applied to the BV samples relative to the H samples. The
orange functions are now statistically significant. Note that this
shifts the midpoint of the housekeeping functions towards the midline.}
\end{figure}

Applying the default scale model of \texttt{gamma=0.5} increases the
dispersion as expected but does little to move the large number of
housekeeping functions toward the midline of no difference. This is as
expected: the mean of the default scale model is based on the CLR
normalization so no shift would be expected over the original ALDEx2
model. Nevertheless, about 50\% of the housekeeping functions are no
longer statistically significantly different. Note that this change is
simple to conduct, has no additional computational complexity and
requires only a slight modification from the analyst.

Up to this point, scale uncertainty has been applied as an extension of
the CLR normalization via the default scale model, but a user-defined
scale adjustment can be applied to each condition, or even each sample
independently through a custom scale matrix. While not necessary in all
datasets (e.g., the transcriptome example discussed in the previous
section and see the Supplementary material), it can greatly improve
modeling in certain cases, especially when the scale is highly
asymmetric between conditions. In order to specify a scale model from
scratch, we need to revisit the concept that all normalizations in
widespread use are actually ratios with the denominator implied by the
normalization. Therefore, we can easily deviate from a certain
normalization (e.g., the geometric mean assumption implied by the CLR)
by specifying a total model based on the mean difference between
conditions. While knowing the mean difference between conditions may
seem cumbersome in practice, it is the \emph{relationship} between the
group scale values that is important, not their raw values (Nixon et
al.~2023). This can be illustrated quite simply by starting with the
mean ratio in \(G\) between groups for the yeast transcriptome dataset
which are \(6.05\times 10^{-5}\) for snf2 and \(5.08\times 10^{-5}\) for
WT; their ratio being 1.17, or a 0.17-fold difference. Using this
information, we can recapitulate the differential abundance analysis in
Figure 2B and 2C exactly by using setting the mean denominator of group
1 to 1, and group 2 to 1.17 with a gamma of 0.5 as shown in Supplemental
Figure 2. This ratio can be adjusted to alter the mean assumption placed
on the group scale values.

With this background, we can understand how to apply a user-defined
scale model to the metratranscriptome dataset. While a user-defined
scale model can be quite flexible with relative scales that are distinct
for each group (or even each sample) along with their uncertainties,
here we focus on using the \texttt{aldex.makeScaleMatrix()} function.
This function uses a logNormal distribution to build a scale matrix
given a user-specified mean difference between groups and uncertainty
level. Applying a per-group relative differential scale of 0.15 moves
the housekeeping functions to the midline of no difference, and applying
a gamma of 0.5 provides the same dispersion as using a gamma of 0.5 with
the CLR (Figure 3C). Note that now a significant number of functions are
differentially up in BV that were formerly classed as not different
without scale, or when only a default scale was applied. These former
false negatives are noted in orange in each panel. Inspection of the
functions shows that these are largely missing from the Lactobacillus
species and so should actually be captured as differentially abundant.
Thus, applying a differential scale allows us to distinguish between
both false positives (housekeeping functions in cyan) and false
negatives (orange functions) even in a very difficult to analyze
dataset. We suggest that the default scale model is sufficient when the
data are approximately centred. However, when datasets are not well
centred or when the investigator has prior information about the
underlying biology, we advocate developing and using a user-specified
scale model.

@@

Biological count data derived from HTS can be decomposed into two parts:
the relative (compositional) and the absolute (scale), and the product
of these generates a fully scaled biological system (Nixon et al.~2023).
Biological systems are both predictably variable and stochastic, and
current measurement methods that rely on high throughput sequencing fail
to capture all of that variation, particularly variation due to scale.
In the absence of information external to the sequencing run itself, no
normalisation method can recapture any of the scale information (Lovén
et al.~2012). Even so, many methods such as flow cytometry (Vandeputte
et al.~2017), spike-in probes and fluorescent in-situ hybridization
(Lovén et al.~2012; Marguerat et al.~2012) have been proposed as a means
to do this. In the context of the updated ALDEx2 software suite, these
methods can be useful if they contain information on the particular
scale the researcher is interested in.

The ALDEx2 R package is readily amenable for incorporating scale
uncertainty. Originally, this tool used the only the Dirichlet
distribution to sample compositional uncertainty. Rather than applying
the CLR normalization as in the original ALDEx2, scale uncertainty can
be added through the use of a scale model with no additional
computational complexity. Scale uncertainty can be sampled from any
distribution depending on prior knowledge or preference. By default,
ALDEx2 samples scale from a logNormal distribution inspired by the CLR
normalization. However, there is the option to introduce a full scale
model that encapsulates both uncertainty and asymmetry in underlying
scale.

All normalizations attempt to make the samples in a dataset commensurate
but do not explicitly address the scale of the underlying system. The
two supplementary tables show that these normalizations can be
interpreted as scale assumptions. However, the general lack of scale
information has important consequences for the analysis of HTS datasets.
One issue is that analysis tools seem over-powered with even moderate
sample sizes, and different tools have different intrinsic power, Type 1
and Type 2 error rates (Schurch et al.~2016) and fail to account for
false discovery rates properly (Hawinkel et al. 2018; Li et al. 2022).
Traditional recommendations would suggest that using small sample sizes
in analysis leads to less reliability and reproducibility in analyses
since surprisingly large sample sizes are needed to determine
reproducible p-values (e.g., Halsey et al. (2015)). However,
recommendations to use small sample sizes are widespread in multivariate
datasets such as RNA-seq datasets. We argue that this is a hallmark of
unacknowledged bias: since normalizations are typically assumed to be
correct with no uncertainty, smaller samples offer an unintentional
buffer against this misspecification. Another issue is that datasets are
difficult to analyze when they contain systematic asymmetry, with
different tools exhibiting differing pathologies with these datasets
(Robinson and Oshlack 2010; Wu et al. 2021).

Many groups have conducted benchmarking studies on different tools and
normalizations used for the analysis of datasets such as transcriptomes
(Bullard et al. 2010; Soneson and Delorenzi 2013; Schurch et al. 2016;
Quinn, Crowley, and Richardson 2018) and microbiomes (Thorsen et al.
2016; Weiss et al. 2017; Hawinkel et al. 2018). Generically, it is
observed that the actual agreement between methods can be modest, and
there is usually a recommendation that some methods are better suited
for individual datasets or for specific types of analysis;
e.g.,\textasciitilde`Normalization and microbial differential abundance
strategies depend upon data characteristics' (Weiss et al. 2017).{]}
However, another interpretation of such studies is that different
normalizations can be better suited to different datasets by
happenstance, but it is not obvious how the use of specific
normalizations could be justified on a per dataset basis. There are now
published examples of multiple large replication efforts in psychology
(Open Science Collaboration 2015), cancer biology (Rodgers and Collings
2021), ecology and evolution (Gould and et.al 2023). In a nutshell,
experiences from these studies indicate that published data typically
have higher effect sizes and appear more significant than replication
efforts with the same data, or with newly generated replication data
(Rodgers and Collings 2021). Thus, most published data should be viewed
with suspicion (Ioannidis 2005) unless efforts have been made to ensure
that the results are robust.
